---
title: "Building your own FLStock and FLIndices"
subtitle: "a4a Training Course — Module 1.5"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## Where we are

You've already build from CSV files:

- an **`FLStock`** — 14 slots filled, 3 left for the model
- an **`FLIndex`** — the survey, with its timing

::: {.callout-tip title="This afternoon"}
The same thing, with **your** data. Then the first assessment.
:::

## The recipe

```
CSV  ->  data.frame  ->  FLQuant  ->  FLStock  ->  checks  ->  sca()
                                  \-> FLIndex -> FLIndices -/
```

Nothing new. The work is in the **details of your files**, not in FLR.

## Your files will not look like ours

Our demo data is **long** — one row per age and year:

```
age, year, data
  1, 1998, 91485.1
```

Most people's data is **wide** — ages down, years across:

```
age, 1998, 1999, 2000, ...
  1, 91485.1, 90945.4, ...
```

Both are fine. Wide needs one extra step:

```{r}
#| eval: false
dat <- read.csv("catch_n.csv", row.names = 1)
flq <- FLQuant(as.matrix(dat), dimnames = list(age = ages, year = yrs))
```

## Decisions only you can make

| Decision | Question |
|:--|:--|
| **Plusgroup** | Which age absorbs everything older? |
| **fbar range** | Which ages average into reported $\bar{F}$? |
| **Year range** | Where do you trust the catch data from? |
| **Units** | Numbers in thousands or units? Weights in kg or g? |

::: {.callout-warning title="These are assessment choices"}
FLR will not question them. Write down what you chose and why —
you will be asked on Friday.
:::

## Useful when reshaping

```{r}
#| eval: false
setPlusGroup(stk, 6)        # collapse ages 6+ into a plusgroup
window(stk, start = 2000)   # cut years, or extend with NA
trim(flq, age = 1:6)        # subset a dimension by name
standardUnits(stk)          # the units FLR expects, slot by slot
```

`setPlusGroup()` sums numbers and takes weighted mean weights —
do not do it by hand.

## When something is missing

| You don't have | Do this |
|:--|:--|
| Discards | `landings = catch`, leave discard slots empty |
| `stock.wt` | Use `catch.wt` as an approximation, and say so |
| Age-varying `M` | One value everywhere: `FLQuant(0.2, ...)` |
| Maturity by year | One ogive, recycled across years |

::: {.notes}
Every one of these is a real assumption. The point is not that they are
forbidden, but that they must be deliberate and written down.
:::


## Check before you fit

```{r}
#| eval: false
validObject(stk)   # does it satisfy the class?
verify(stk)        # negative weights? maturity > 1? unit mismatches?
```

::: {.callout-warning title="verify() will always flag three things"}
`anyna`, `harvest` and `cohorts` fail on **any** unfitted stock —
`stock`, `stock.n` and `harvest` are empty on purpose. Ignore those three;
read everything else.
:::

Then spot-check **one value** per file. A transposed matrix gives a
plausible, completely wrong assessment.

## Then, the first fit

```{r}
#| eval: false
idxs <- FLIndices(MEDITS = idx)

fit <- sca(stock = stk, indices = idxs)

stk.fit <- stk + fit          # the estimated slots, filled in

plot(FLStocks(input = stk, assessment = stk.fit))
```

`stk + fit` puts `stock.n`, `stock` and `harvest` back into the object.

::: {.notes}
This is the Day 1 payoff. Do not interpret the fit — that is Day 2.
Today the point is only that the data are in a shape the model accepts.
:::

## Summary

- The pipeline is always the same; only your files differ
- **Wide or long** — both work, wide needs `as.matrix()`
- Plusgroup, fbar range, year range and units are **your** decisions
- Check that you have filled all necessary slots
- `stk + fit` returns the assessed stock - fills the 3 missing ones

Next: **Module 1.6 — inspecting and visualising what you have built.**

## Your turn

::: {.callout-tip title="Exercise"}
1. Run `handson_M1.5_build_from_data.R` — the demo stock, all the way
   to a first `sca()` fit
2. Then open `exercise_M1.5_own_data.R` and work through it with
   **your own files**
3. The script ends with a **PASS/FAIL checklist**. Get it all green
   before you fit - or be sure why they fail.
:::

