---
title: "FLIndex — survey and CPUE data"
subtitle: "a4a Training Course — Module 1.4"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
library(ggplotFL)

# Build the demo index quietly, exactly as the hands-on script does.
dd <- "../../_data/MUT_GSA32"
idx.df  <- read.csv(file.path(dd, "index.csv"))
idx.flq <- FLQuant(idx.df$data,
                   dimnames = list(age = 1:10, year = 1998:2017))

idx <- FLIndex(idx.flq)
index(idx) <- idx.flq
range(idx)["startf"] <- 6 / 12
range(idx)["endf"]   <- 8 / 12
name(idx) <- "MEDITS"
desc(idx) <- "Medits in GSA 32"
```

## Why we need an index at all

Catch data alone cannot tell you whether a small catch means

- **few fish**, or
- **little fishing**

::: {.callout-warning title="The identification problem"}
Catch $= F \times N$. One equation, two unknowns.

A survey gives an independent look at $N$ — that is what breaks the tie.
:::

::: {.notes}
This is the single most important idea in the module. If they take nothing
else away: the index is what makes the assessment identifiable.
:::

## What an FLIndex holds

Twelve slots — but you normally fill **very few**.

| Slot | What it is |
|:--|:--|
| `index` | the abundance series, ages $\times$ years |
| `index.var` | variance / weighting, optional |
| `range` | ages, years, and **survey timing** |
| `name`, `desc` | which survey is this |
| `type`, `distribution` | metadata, rarely changed |
| `catch.n`, `catch.wt`, `effort`, `sel.pattern`, `index.q` | usually left empty |

For a standard a4a fit: **`index` plus `range`**.

## The index is relative, not absolute

The survey does not count every fish. It sees a **fraction** of them:

$$I_{a,y} = q_a \cdot N_{a,y}$$

- $q$ is **catchability** — the fraction the survey sees
- The model estimates $q$; you only supply $I$

::: {.callout-tip title="Which means"}
The **units do not matter** and the absolute level does not matter.
What matters is how the index changes **through time**.
:::

::: {.notes}
Say plainly: do not rescale your index, do not worry that it is "too small".
Trends are the signal. We configure the q submodel on Day 2 and Day 3.
:::

## `range` — and the two numbers people forget

```{r}
range(idx)[c("startf", "endf")]
```

The survey's start and end as a **fraction of the year** —
June to August is `6/12` to `8/12`.

::: {.callout-warning title="Why it matters"}
They tell the model **how much mortality has already happened** when the
survey sees the fish. Set them wrong and the index is compared against the
population at the wrong moment.
:::

## Age-structured or biomass?

| Class | Index is | Use when |
|:--|:--|:--|
| `FLIndex` | numbers **at age** | your survey is aged |
| `FLIndexBiomass` | a **single** biomass series | no age data, e.g. acoustic |

Same idea, same `range`, same survey timing.
`FLIndexBiomass` simply has one row instead of ten.

## One index or many

`sca()` takes **`FLIndices`** or an `FLIndex`:

```{r}
#| eval: false
idx  <- FLIndex(idx.flq)      # one survey

# you can add more surveys
idxs <- FLIndices(idx1 = idx, idx2 = idx)

fit <- sca(stock = stk, indices = idxs)
```

::: {.callout-warning title="Some methods still require FLIndices()"}
The safest option is to wrap even a single index in an `FLIndices` object.
:::


## Looking at it

```{r}
#| fig-height: 2.6
plot(idx) + ggtitle(name(idx))
```

## Pitfalls worth knowing

- **Zeros and NAs** are not the same. A true zero is information;
  a missing survey year should be `NA`
- **Do not rescale** to match catch — the model estimates $q$ for that
- **One survey per `FLIndex`.** Two surveys means two objects in `FLIndices`
- If the survey changed vessel or gear, you need to think if you need to split to two surveys

## Summary

- The index is what makes $F$ and $N$ separable
- You normally fill only **`index`** and **`range`**
- It is **relative**: trends matter, absolute level does not
- `startf`/`endf` place the survey inside the year — set them


Next: **Module 1.5 — putting stock and index together, and the first fit.**

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M1.4_flindex.R` and build the MEDITS index for MUT GSA 32.

1. Read `index.csv` and build the `FLQuant`
2. Create the `FLIndex` and fill the `index` slot
3. Set the survey timing, `name` and `desc`
4. Plot it, and check its dimensions against the stock
:::


