---
title: "Adding year effects"
subtitle: "a4a Training Course — Module 2.4"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(lattice)
library(latticeExtra)
lattice.options(default.args = list(as.table = TRUE))

data(mut09)
data(mut09.idx)
stk <- mut09
idx <- mut09.idx

f <- function(...) sca(stk, idx, ...)

# where Module 2.3 left off: both age effects, no year effect
fit04 <- f(fmodel = ~factor(age),                qmodel = list(~factor(age)),
           srmodel = ~1, vmodel = list(~1, ~1),  n1model = ~1)
fit05 <- f(fmodel = ~factor(age) + factor(year), qmodel = list(~1),
           srmodel = ~1, vmodel = list(~1, ~1),  n1model = ~1)
fit06 <- f(fmodel = ~factor(age) + factor(year), qmodel = list(~factor(age)),
           srmodel = ~1, vmodel = list(~1, ~1),  n1model = ~factor(age))
fit07 <- f(fmodel = ~factor(age) + factor(year), qmodel = list(~factor(age)),
           srmodel = ~factor(year), vmodel = list(~1, ~1), n1model = ~factor(age))

res05 <- residuals(fit05, stk, idx)
res06 <- residuals(fit06, stk, idx)
res07 <- residuals(fit07, stk, idx)
```

## A year effect on F

The last set of residuals showed a clear pattern **across years** — more
positive early in the series, more negative in the recent years. As for age,
we answer it with a `factor`.

```{r}
#| eval: false
fit05 <- sca(stk, idx,
             fmodel = ~factor(age) + factor(year),   # <- year added
             qmodel = list(~1),
             srmodel = ~1, vmodel = list(~1, ~1), n1model = ~1)
```

::: {.callout-tip title="And the scale is pinned down"}
```{r}
#| echo: false
round(range(c(fbar(stk + fit05))), 2)
```
$\bar{F}$ is a real number again, not $10^{-15}$. Letting $F$ move through
time is what fixes the degeneracy from Module 2.3.
:::

## Year effect on F — residuals by year

```{r}
#| echo: false
#| fig-height: 3.0
plot(res05)
```

## Year effect on F — residuals by age

```{r}
#| echo: false
#| fig-height: 2.3
plot(res05, auxline = "l", by = "age")
```

::: {.callout-note title="Not finished"}
Much less pattern than before — but the survey catchability still looks like
it has a year trend, and recruitment, $N_1$ and the variances are all still
constants.
:::

## Why we do **not** add year effects to q

A well-designed scientific survey should see abundance change because
**abundance changed**.

::: {.callout-warning title="Modelling a year trend in q"}
…attributes that change to the **survey design** instead, and removes it
from the abundance. Think carefully before you do it.
:::

The exception is a **commercial CPUE**, where changes in selectivity or
fishing behaviour really can enter the index — although standardisation
should have dealt with the technical part.

The formulas are the same ones: period breaks, trends, and so on.

## The initial population, N1

The first year has no earlier data to reconstruct its cohorts from, so
`n1model` sets the numbers at age in year 1.

```{r}
#| eval: false
fit06 <- sca(stk, idx,
             fmodel  = ~factor(age) + factor(year),
             qmodel  = list(~factor(age)),
             srmodel = ~1, vmodel = list(~1, ~1),
             # <- was ~1
             n1model = ~factor(age))
```

It affects the **lower triangle** of the population matrix — the early years
— and the catches there. So zoom into the start of the series to see it.

## N1 as an intercept: 2003–2010 by age

```{r}
#| echo: false
#| fig-height: 3.0
plot(window(res05, end = 2010), auxline = "l", by = "age")
```

## N1 with an age effect: 2003–2010 by age

```{r}
#| echo: false
#| fig-height: 3.0
plot(window(res06, end = 2010), auxline = "l", by = "age")
```

Catch and survey residuals both improve considerably. The following years
improve too, to different degrees.

## The stock-recruitment submodel {.shrink}

Constant recruitment is a strong assumption. Let it vary over time:

```{r}
#| eval: false
fit07 <- sca(stk, idx,
             fmodel  = ~factor(age) + factor(year),
             qmodel  = list(~factor(age)),
             # <- was ~1
             srmodel = ~factor(year),
             vmodel  = list(~1, ~1), n1model = ~factor(age))
```

::: {.callout-tip title="Why it matters everywhere"}
Unmodelled recruitment variability does not stay put: it was being picked up
as a **trend in the survey catchability**, and cohort dynamics carried it
into the other ages.
:::

## Recruitment varying — residuals by year

```{r}
#| echo: false
#| fig-height: 3.0
plot(res07)
```

The year trends present in the previous models are gone.

## Summary

- A year effect on $F$ **breaks the scale degeneracy** from Module 2.3
- Do **not** put year trends in $q$ for a scientific survey — think first
- `n1model` fixes the **early years**, propagating along cohorts
- `srmodel` lets recruitment vary; unmodelled, it contaminates $q$ and the catch
- Each step was driven by **what the residuals showed**, one change at a time

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M2.4_year_effects.R`.

1. Add the year effect and check `fbar` is sensible again
2. Add `n1model`, and compare the early years before and after
3. Add `srmodel`, and watch the year trends leave the survey residuals
:::
