---
title: "Module Title"
subtitle: "a4a Training Course"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(ggplot2)
theme_set(theme_bw())
```

## Outline

- Point 1
- Point 2
- Point 3

::: {.notes}
Speaker notes go here. These are not shown on the slide.
:::

## Main Content Slide

Content here.

### With a sub-heading

More content.

## Code Example

```{r}
#| label: example-code
#| echo: true
# Example R code
data(ple4)
plot(ple4)
```

## Key Message Slide

::: {.callout-warning title="Key message"}
**a4a has no process error** — residuals are the main tool for diagnosing model
fit and guiding submodel configuration.
:::

## Exercise Slide

::: {.callout-tip title="Exercise"}
1. Load your own data
2. Create an FLStock object
3. Check the dimensions and units
4. Plot the stock
:::

## Optional Geeky Stuff

::: {.callout-note title="Optional Geeky Stuff"}
Detailed mathematical content goes here.

$$\mathbf{y} = \beta_0 + \beta_1\mathbf{x} + \epsilon$$

Keep these slides at the end of the module so they can be shown or skipped
without breaking the flow.
:::

## Two-Column Layout

::: {.columns}

::: {.column width="50%"}
**Left column**

- Item 1
- Item 2
:::

::: {.column width="50%"}
```{r}
#| eval: false
# Code on the right
sca(stk, idx)
```
:::

:::

## Summary

- Takeaway 1
- Takeaway 2
- Takeaway 3

## Hands-on Session

Now it's your turn! Work on your own data.

**Objectives for this session:**

1. Objective 1
2. Objective 2
3. Objective 3

Come back for questions at any time.
