---
title: "FLQuant — the building block"
subtitle: "a4a Training Course — Module 1.2"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
```

## The one class to understand

All data in FLR live inside an **`FLQuant`**.

- Catch numbers at age
- Weights at age
- Natural mortality, maturity
- Survey indices

::: {.callout-tip title="Why we start here"}
`FLQuant` is not one more class to learn — it is **the** class.
Everything we build for the rest of the week is made of these.
:::

::: {.notes}
Twenty minutes on this pays for itself all week. If the shape of an FLQuant is
clear, the objects built from them stop being mysterious.
:::

## An FLQuant is always six-dimensional

| Dimension | What it holds | Usual value |
|:--|:--|:--|
| `quant` | age, length, … | `age` |
| `year` | time | your year range |
| `unit` | sex or stock component | `unique` |
| `season` | within-year structure | `all` |
| `area` | spatial strata | `unique` |
| `iter` | iterations | `1` |

Most of the time you only care about the **first two** and when we visit `simulate()` we will inspect the `iter`

## Even an empty one has six dimensions

```{r}
dim(FLQuant())
```

The first dimension is the flexible one — hence the name `quant`:

```{r}
names(dimnames(FLQuant(dimnames = list(length = c(10, 20, 30)))))
```

::: {.notes}
This is the single most common source of confusion: people expect a 2D matrix
and get six dimensions. Say clearly that four of them are usually degenerate.
:::

## Creating one — from a vector {.shrink}

Values fill **down the quant first**, then across years.

```{r}
FLQuant(1:6, dimnames = list(age = 1:3, year = 2000:2001), units = "kg")
```

## Creating one — other ways

```{r}
#| eval: false
FLQuant()                                       # empty, 1x1x1x1x1x1
FLQuant(0.2, dimnames = list(age = 1:2))        # one value, recycled
FLQuant(matrix(1:6, nrow = 3), quant = "age")   # from a matrix
FLQuant(array(rlnorm(100), dim = c(10, 5, 2)))  # from an array
as.FLQuant(df)                                  # from a data.frame
```

::: {.callout-tip title="The usual route from your own data"}
`read.csv()` → `as.matrix()` → `FLQuant()`

That is exactly what we do in Module 1.5 with your files.
:::

## Data and metadata travel together

```{r}
flq <- FLQuant(1:6, 
    dimnames = list(age = 1:3, year = 2000:2001), 
    units = "kg")
dim(flq)
dimnames(flq)$year
units(flq)
```


## FLQuant() can handle (most of the times) units:

```{r}
n <- FLQuant(1000, units = "1000")   # numbers, in thousands
w <- FLQuant(0.5,  units = "kg")     # mean weight

units(n * w)
```

FLR does the unit algebra for you: thousands $\times$ kg $=$ **tonnes**.

::: {.callout-warning title="Which means"}
If your units are wrong, the arithmetic is still wrong —
it just fails **quietly**. Set them when you build the object.
:::

## Two things that catch everybody

**1. Dimnames are character, not numbers**

```{r}
dimnames(flq)$year
```

So subset by label, `flq[, "2000"]`, or by position, `flq[, 1]` —
but `flq[, 2000]` is out of bounds.

**2. Dimensions are not dropped by default**

`flq[1, ]` is still a full 6D `FLQuant`. Use `drop = TRUE` to get a plain vector.

## The FLQuant toolset

\begin{center}
\includegraphics[keepaspectratio, height=0.88\textheight]{flquant.png}
\end{center}

## Summary

- One class, **six dimensions** — you mostly use the first two
- `quant` is flexible: `age`, `length`, whatever you need
- Carries **dimnames and units**, not just numbers
- Units are checked and combined by FLR — set them correctly
- Dimnames are **character**; dimensions are **not dropped**

Next: **Module 1.3 — `FLStock`, built entirely out of `FLQuant`s.**

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M1.2_flquant.R` and work through it.

1. Create `FLQuant`s from a value, a vector and a matrix
2. Inspect `dim`, `dimnames` and `units`
3. Arithmetic: what conforms, what does not, and when to use `%*%`
4. Summaries: `quantSums`, `yearMeans`, geometric mean
5. Reproduce a target object from its printed output
:::

## Optional Geeky Stuff

::: {.callout-note title="Why six dimensions?"}
An `FLQuant` extends R's `array`, with a fixed dimension order:

`quant` $\times$ `year` $\times$ `unit` $\times$ `season` $\times$ `area` $\times$ `iter`

Fixing the order is what lets every FLR package agree on what it is
being handed — the "common currency" from Module 1.1.

The 6th dimension, `iter`, stays at 1 all week until **Day 4**, where uncertainty is propagated through it.
:::
