---
title: "Introduction to FLR"
subtitle: "a4a Training Course — Module 1.1"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## The problem FLR was built to solve

Every stock assessment group used to write its own software.

- Methods that could not be compared
- Results that could not be reproduced
- Code that died with the person who wrote it

::: {.callout-warning title="The consequence"}
You could not test whether a management strategy was **robust**,
because you could not connect the pieces.
:::

::: {.notes}
This is the framing of the 2007 paper. Do not dwell — one minute. The experts
have lived this problem themselves, so ask if it sounds familiar.
:::

## What FLR is

*"A development effort directed towards the evaluation of fisheries
management strategies. The overall goal is to develop a **common framework**
to facilitate collaboration within and across disciplines."*

**FLR** — *Fisheries Library for R*

- A **collection of R packages**, not a single program
- A **common set of objects** every package understands
- An **open source** project, still actively developed

\vfill
\tiny Kell *et al.* (2007) *FLR: an open-source framework for the evaluation and
development of management strategies.* ICES J. Mar. Sci. 64(4): 640–646.

## The design principles

| Principle | What it means for you |
|:--|:--|
| **Common currency** | Every package speaks the same objects |
| **Object-oriented** | Data and metadata travel together |
| **Extensible** | New methods plug into the same structures |
| **Open source** | You can read, check and change the code |

::: {.notes}
"Data and metadata travel together" is the one that matters on Day 1 — units
and dimension names are carried by the object, not kept in your head.
:::

## Why this helps you in practice

- Your stock object works with **any** assessment method in FLR
- Swap `a4a` for `XSA` or `SAM` — **the data does not change**
- The output of the assessment feeds **straight into** reference points
  and forecasts, with no re-formatting
- Plotting, summarising and subsetting work the **same way** everywhere

::: {.callout-tip title="The practical payoff"}
You prepare your data **once**, on Day 1.
Everything else this week reuses that same object.
:::

## The FLR universe

| Area | Packages |
|:--|:--|
| Core objects & plotting | `FLCore`, `ggplotFL` |
| Stock assessment | `FLa4a`, `FLXSA`, `FLSAM` |
| Reference points & advice | `FLBRP` |
| Forecasting & projection | `FLasher`, `FLash` |
| Life history & biology | `FLife` |
| Management Strategy Evaluation (MSE) | `mse`, `mseviz`, `FLFishery` |
| Diagnostics | `FLa4a`, `a4adiags` |

You do **not** need to know these. You need to know the six we use.

## The six we use this week

| Package | What it gives us | When |
|:--|:--|:--|
| `FLCore` | The objects: `FLQuant`, `FLStock`, `FLIndex` | Day 1 |
| `ggplotFL` | Plot methods for FLR objects | Day 1 |
| `FLa4a` | The assessment model itself, `sca()` | Days 2–4 |
| `a4adiags` | Retrospective & hindcast diagnostics | Day 4 |
| `FLBRP` | Reference points | Day 5 |
| `FLasher` | Short-term forecasts | Day 5 |

These are exactly the packages your setup check just loaded.

## Two habits worth picking up now

**Use accessors**

```{r}
#| eval: false
catch(stk)            # yes
stk@catch             # no
catch(stk) <- newdata # checks the object stays valid
```

**Every class has a plural** — `FLStock` → `FLStocks`,
`FLIndex` → `FLIndices`, `FLQuant` → `FLQuants`

::: {.notes}
The plural point matters because several FLR methods only accept the plural
class. sca() happens to take either an FLIndex or an FLIndices, but wrapping
a single survey in FLIndices() is the habit that travels safely.
:::

## Getting help

```{r}
#| eval: false
help("FLStock")            # class documentation
?sca                       # a method
methods(class = "FLStock")  # what can I do with this?
showClass("FLStock")
```

- FLa4a documentation: <https://flrproject.org/FLa4a/>
- And: we are around all week

## Looking inside a method

```{r}
#| eval: false
methods(class = "FLStock")   # what can I do with this object?
showMethods("plot")          # which classes does plot() handle?
```

`plot(ple4)` dispatches on **two** arguments — `x="FLStock"`, `y="missing"`:

```{r}
#| eval: false
getMethod("plot", signature = list("FLStock", "missing"))
```

```r
.local <- function (x, metrics = list(Rec = rec, SSB = ssb,
                     Catch = catch, F = fbar), na.rm = TRUE, ...)
```

That default `metrics` list is **why** `plot(ple4)` gives you those four
panels — and how you ask for different ones.

::: {.notes}
Run this live. The full method is ~50 lines, so scroll to the .local line and
stop there — the point is the metrics default, not the whole body. Nothing in
FLR is hidden: if you wonder what a method does, you can always read it.
This also sets up M1.6, where we change metrics to plot what we want.
:::

## Optional Geeky Stuff

::: {.callout-note title="Why S4 and not plain R?"}
FLR uses **S4** classes rather than ordinary lists.

- **Slots are typed** — an `FLStock` cannot hold the wrong thing
- **Validity is checked** on creation and on replacement
- **Methods dispatch on class** — `plot()` knows what it was handed

The cost is a steeper learning curve. The benefit is that a malformed
stock object fails loudly on Day 1, rather than silently producing a
wrong assessment on Day 3.
:::

## Summary

- FLR is a **common framework**, not one program
- Its value is **interoperability**: prepare data once, reuse all week
- We use **six** packages; `FLCore` is the foundation
- Use **accessors**, and watch the **plurals**

Next: **Module 1.2 — `FLQuant`, the building block everything is made of.**
