---
title: "FLStock — structure, slots and meaning"
subtitle: "a4a Training Course — Module 1.3"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)

# Build the demo stock quietly so the slides can show real output.
# This is exactly what the hands-on script does, step by step.
dd <- "../../_data/MUT_GSA32"
yrs <- 1998:2017
ages <- 1:10
fq <- function(x, units = "") FLQuant(x, dimnames = list(age = ages, year = yrs),
                                      units = units)

land    <- read.csv(file.path(dd, "landings.csv"))
land.n  <- read.csv(file.path(dd, "landings_n.csv"))
land.wt <- read.csv(file.path(dd, "landings_wt.csv"))
disc    <- read.csv(file.path(dd, "discards.csv"))
disc.n  <- read.csv(file.path(dd, "discards_n.csv"))
disc.wt <- read.csv(file.path(dd, "discards_wt.csv"))
stk.wt  <- read.csv(file.path(dd, "stock_wt.csv"))

landings.flq    <- FLQuant(land$data, dimnames = list(age = "all", year = yrs), units = "t")
landings.n.flq  <- fq(land.n$data,  "1000")
landings.wt.flq <- fq(land.wt$data, "kg")
discards.flq    <- FLQuant(disc$data, dimnames = list(age = "all", year = yrs), units = "t")
discards.n.flq  <- fq(disc.n$data,  "1000")
discards.wt.flq <- fq(disc.wt$data, "kg")
stock.wt.flq    <- fq(stk.wt$data,  "kg")

catch.flq   <- landings.flq + discards.flq
catch.n.flq <- landings.n.flq + discards.n.flq
catch.wt.flq <- (discards.n.flq / catch.n.flq) * discards.wt.flq +
                (landings.n.flq / catch.n.flq) * landings.wt.flq

m.flq      <- fq(0.1)
mat.flq    <- fq(rep(c(0, 0.5, 0.5, rep(1, 7)), times = length(yrs)))
m.spwn.flq <- fq(0)
h.spwn.flq <- fq(0)

stk <- FLStock(catch.n.flq)
catch(stk) <- catch.flq;       catch.n(stk) <- catch.n.flq
catch.wt(stk) <- catch.wt.flq
landings(stk) <- landings.flq; landings.n(stk) <- landings.n.flq
landings.wt(stk) <- landings.wt.flq
discards(stk) <- discards.flq; discards.n(stk) <- discards.n.flq
discards.wt(stk) <- discards.wt.flq
stock.wt(stk) <- stock.wt.flq
m(stk) <- m.flq; mat(stk) <- mat.flq
m.spwn(stk) <- m.spwn.flq; harvest.spwn(stk) <- h.spwn.flq
range(stk)["minfbar"] <- 2
range(stk)["maxfbar"] <- 6
name(stk) <- "MUT"
desc(stk) <- "MUT in GSA32. WGSAD 2018"
units(harvest(stk)) <- "f"
```

## What an FLStock is

An **`FLStock`** is a container holding your whole perception of the stock.

- **17 `FLQuant` slots** — all sharing the same ages and years
- plus `name`, `desc` and `range`

::: {.callout-tip title="Nothing new to learn"}
Every slot is an `FLQuant`, exactly as in Module 1.2.
An `FLStock` is just a labelled box of them.
:::

::: {.notes}
Stress the "same shape" point: if one slot has different dimensions from the
rest, the object will not validate. That is a feature, not an obstacle.
:::

## The 20 slots, grouped

| Group | Slots |
|:--|:--|
| Catch | `catch`, `catch.n`, `catch.wt` |
| Landings | `landings`, `landings.n`, `landings.wt` |
| Discards | `discards`, `discards.n`, `discards.wt` |
| Biology | `m`, `mat`, `m.spwn`, `harvest.spwn` |
| Population | `stock`, `stock.n`, `stock.wt` |
| Exploitation | `harvest` |
| Metadata | `name`, `desc`, `range` |

## The pattern: total, numbers, weight

Catch, landings and discards each come as **three** slots:

| Slot | Meaning | Dimensions | Units |
|:--|:--|:--|:--|
| `catch` | total weight landed + discarded | 1 $\times$ years | `t` |
| `catch.n` | numbers **at age** | ages $\times$ years | `1000` |
| `catch.wt` | mean weight **at age** | ages $\times$ years | `kg` |

Note the first one is a **single row** (`age = "all"`), not a matrix.

::: {.notes}
This trips people up: catch is a time series, catch.n and catch.wt are
matrices. Same for landings and discards.
:::

## Biology — what you must supply

| Slot | Meaning | Typical |
|:--|:--|:--|
| `m` | natural mortality at age | assumed, e.g. 0.1 |
| `mat` | proportion mature at age | 0 → 1 ogive |
| `m.spwn` | fraction of **M** before spawning | 0 |
| `harvest.spwn` | fraction of **F** before spawning | 0 |

`m.spwn` and `harvest.spwn` say *how far into the year* spawning happens.
Both 0 means spawning on 1 January.

## Population and exploitation — what the model gives back

| Slot | Meaning | Who fills it |
|:--|:--|:--|
| `stock.wt` | mean weight at age in the stock | **you** |
| `stock.n` | numbers at age in the stock | the model |
| `stock` | total stock biomass | the model |
| `harvest` | fishing mortality at age | the model |

::: {.callout-warning title="The key distinction"}
You supply **14** of the 17 `FLQuant` slots.
`stock`, `stock.n` and `harvest` are what `sca()` **estimates** —
leave them empty.
:::

## `range` — seven numbers that matter

```{r}
range(stk)
```

- `min`, `max`, `plusgroup` — the age structure
- `minyear`, `maxyear` — the time span
- `minfbar`, `maxfbar` — **which ages are averaged** to report $\bar{F}$

::: {.notes}
minfbar/maxfbar are a choice, not data. They change the F you report without
changing the fit at all. Ask them what their assessment uses.
:::

## Metadata is not decoration

```{r}
name(stk)
desc(stk)
```

Six months from now, this is how you know which run you are looking at.

## Consistency: the model trusts you

Nothing checks that your slots agree with each other. You must ensure:

```{r}
#| eval: false
catch.n  == landings.n  + discards.n
catch    == landings    + discards
```

And `catch.wt` is the **catch-weighted** mean of the two:

```{r}
#| eval: false
catch.wt <- (discards.n / catch.n) * discards.wt +
            (landings.n / catch.n) * landings.wt
```

::: {.callout-warning title="Not a plain average"}
Averaging `landings.wt` and `discards.wt` directly is wrong —
discards are small fish, and there are usually many more of them.
:::

## A trap worth knowing

`FLQuant` matches `dimnames` **by name**, not by position:

```{r}
#| eval: false
# year written first, but you still get ages x years
FLQuant(x, dimnames = list(year = 1998:2017, age = 1:10))
```

That works. But the **values** fill in array order — quant first.

::: {.callout-warning title="So"}
This only gives the right answer because the CSV is sorted **age within year**.
A year-major file produces a silently **transposed** matrix, with no error.

Always spot-check one value: `flq["3", "2005"]` against the raw file.
:::

## Checking your object

```{r}
#| eval: false
summary(stk)      # dimensions and ranges, slot by slot
plot(stk)         # the four-panel overview
units(stk)        # are the units what you think?
validObject(stk)  # does it satisfy the class definition?
verify(stk)       # FLCore's sanity checks
```

::: {.notes}
verify() is the one they will not know about. It catches negative catches,
maturity outside 0-1, and similar. Demonstrate it live on Day 1 if there is
time, otherwise it reappears in M1.5 with their own data.
:::

## Summary

- An `FLStock` is **17 `FLQuant`s** plus `name`, `desc`, `range`
- Totals are `1 x years`; numbers and weights are `ages x years`
- You supply **14**; `stock`, `stock.n`, `harvest` come from the model
- `range` carries the **fbar ages** — a choice you must make
- Nothing enforces internal consistency: **that is your job**

Next: **Module 1.4 — `FLIndex`, the survey side of the data.**

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M1.3_flstock.R` and build the MUT GSA32 stock from CSV files.

1. Read each file and turn it into an `FLQuant`
2. Derive `catch`, `catch.n` and `catch.wt` from landings and discards
3. Fill the biology slots
4. Assemble the `FLStock`, then set `range`, `name`, `desc` and `units`
5. Plot it
:::

## Optional Geeky Stuff

::: {.callout-note title="Why a class and not a list?"}
`FLStock` is an S4 class with a **validity function**. Every assignment
through an accessor re-checks it:

- all `FLQuant` slots must share the same `year` and `iter` dimensions
- the quant dimension must match, except for the `1 x years` totals

So `catch.n(stk) <- wrong_shape` fails **immediately**, at the line that
caused it — rather than fifty lines later inside `sca()`, where the error
message would tell you nothing useful.
:::
