---
title: "Inspecting, manipulating and visualising"
subtitle: "a4a Training Course — Module 1.6"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(patchwork)
theme_set(theme_bw())

# The MUT stock you built this morning, fitted, so the estimated slots
# are populated and the plots have something to show.
dd <- "../../_data/MUT_GSA32"
yrs <- 1998:2017; ages <- 1:10
rd <- function(f) read.csv(file.path(dd, f))
fq <- function(x, u = "") FLQuant(x, dimnames = list(age = ages, year = yrs), units = u)
tot <- function(x) FLQuant(x, dimnames = list(age = "all", year = yrs), units = "t")

ln <- fq(rd("landings_n.csv")$data, "1000"); lw <- fq(rd("landings_wt.csv")$data, "kg")
dn <- fq(rd("discards_n.csv")$data, "1000"); dw <- fq(rd("discards_wt.csv")$data, "kg")
cn <- ln + dn
stk <- FLStock(cn)
catch.n(stk) <- cn; catch(stk) <- tot(rd("landings.csv")$data) + tot(rd("discards.csv")$data)
catch.wt(stk) <- (dn / cn) * dw + (ln / cn) * lw
landings(stk) <- tot(rd("landings.csv")$data); landings.n(stk) <- ln; landings.wt(stk) <- lw
discards(stk) <- tot(rd("discards.csv")$data); discards.n(stk) <- dn; discards.wt(stk) <- dw
stock.wt(stk) <- fq(rd("stock_wt.csv")$data, "kg")
m(stk) <- fq(0.1, "m"); mat(stk) <- fq(c(0, 0.5, 0.5, rep(1, 7)))
m.spwn(stk) <- fq(0); harvest.spwn(stk) <- fq(0)
range(stk)[c("minfbar", "maxfbar")] <- c(2, 6)
name(stk) <- "MUT"; units(harvest(stk)) <- "f"

idx <- FLIndex(fq(rd("index.csv")$data))
index(idx) <- fq(rd("index.csv")$data)
range(idx)[c("startf", "endf")] <- c(6/12, 8/12)
name(idx) <- "MEDITS"

fit <- sca(stk, FLIndices(MEDITS = idx))
stk.fit <- stk + fit
```

## Three ways to look at an object

| Way | Use it for |
|:--|:--|
| `summary()`, `print()` | dimensions, ranges, units — the quick check |
| `plot()` | the standard overview, one line |
| `as.data.frame()` $\to$ `ggplot()` | anything you want, exactly how you want it |

The first two are free. The third is where you spend your time when
building a report.

## `metrics()` — the numbers people ask for

```{r}
names(metrics(stk.fit))
```

Each is a method you can also call directly:

```{r}
#| eval: false
ssb(stk)   rec(stk)   fbar(stk)   catch(stk)
tsb(stk)   vb(stk)    fapex(stk)  sp(stk)
```

::: {.notes}
sp() is surplus production — useful on Day 5 when we talk about production
and reference points. Do not dwell here.
:::

## `plot()` is a method, not a function

One name, different behaviour per class:

```{r}
#| eval: false
plot(flq)          # an FLQuant   — a single time series
plot(stk)          # an FLStock   — the four-panel overview
plot(idx)          # an FLIndex   — index by age
plot(FLStocks(...)) # several stocks — overlaid for comparison
```

Remember Module 1.1: `getMethod("plot", list("FLStock", "missing"))`
shows you exactly what it does.

## The four-panel default

```{r}
#| fig-height: 2.7
plot(stk.fit)
```

## Change what it shows

The default came from that method's `metrics` argument — so override it:

```{r}
#| fig-height: 1.9
plot(stk.fit, metrics = list(SSB = ssb, `bar(F)` = fbar))
```

::: {.callout-warning title="Panel names are plotmath — no spaces"}
`Mean F` errors. `Fbar` works. `` `bar(F)` `` renders as $\bar{F}$.
:::

## Comparing runs

`FLStocks` is how you put two assessments side by side:

```{r}
#| fig-height: 2.5
plot(FLStocks(input = stk, fitted = stk.fit))
```

::: {.notes}
This is the plot they will use all week: base run vs new configuration.
:::

## The escape hatch: `as.data.frame()`

Every FLR object converts to a long data.frame:

```{r}
head(as.data.frame(catch.n(stk)), 3)
```

From here it is ordinary `ggplot2` — `data`, `year`, `age` are just columns.

## Bubble plots {.shrink}

Size by value, split by age and year:

```{r}
#| fig-height: 1.8
ggplot(catch.n(stk), aes(year, factor(age), size = data)) +
  geom_point(shape = 21) + labs(y = "age", size = "catch.n")
```



## Cohorts

```{r}
#| fig-height: 2.2
ggplot(stock.n(stk.fit), aes(year, data, colour = factor(age))) +
  geom_line() + scale_y_log10() + labs(colour = "age") +
  guides(colour = guide_legend(ncol = 2))
```

Following a strong year class is a quick sanity check on the assessment.

## Assembling figures

`patchwork` combines any ggplots with `+`, `/` and `|`:

```{r}
#| eval: false
p1 <- plot(ssb(stk.fit)) + ggtitle("SSB")
p2 <- plot(fbar(stk.fit)) + ggtitle("Mean F")

(p1 | p2) / plot(rec(stk.fit))
```

Useful on Day 5, when everything has to fit on one page.

## Summary

- `summary()` for the check, `plot()` for the overview
- `metrics()` gives the standard quantities; each is also a method
- `plot()` takes a **`metrics` argument** — you choose the panels
- `FLStocks()` compares runs; you will use it constantly
- `as.data.frame()` then `ggplot2` for everything else

::: {.callout-tip title="That is Day 1"}
Your data is in FLR, checked, and fitted.
Tomorrow we start asking whether the fit is any good.
:::

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M1.6_visualisation.R`.

1. `summary`, `units`, `metrics` and the individual methods
2. Subset and trim: `window()`, `trim()`, `[`
3. `plot()` per class, change its `metrics`, then **write your own metric**
4. `FLQuants`, faceting, log scales, and a 3D `wireframe`
5. `as.data.frame()` → bubbles, boxplots, cohorts
6. `geom_flpar` for reference points, and `cohcorrplot`
7. Assemble with `patchwork` and save to file
:::
