---
title: "Adding age effects"
subtitle: "a4a Training Course — Module 2.3"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(lattice)
library(latticeExtra)
lattice.options(default.args = list(as.table = TRUE))

data(mut09)
data(mut09.idx)
stk <- mut09
idx <- mut09.idx

ff <- function(...) sca(stk, idx, srmodel = ~1, vmodel = list(~1, ~1),
                        n1model = ~1, ...)

fit01 <- ff(fmodel = ~1,           qmodel = list(~1))
fit02 <- ff(fmodel = ~factor(age), qmodel = list(~1))
fit03 <- ff(fmodel = ~1,           qmodel = list(~factor(age)))
fit04 <- ff(fmodel = ~factor(age), qmodel = list(~factor(age)))

res02 <- residuals(fit02, stk, idx)
res03 <- residuals(fit03, stk, idx)
res04 <- residuals(fit04, stk, idx)
```

## The signal we are answering

The mean model left a clear **age** pattern in both data sources.

::: {.callout-tip title="The method"}
Explore the age effect on $F$ and on $q$ **in isolation**, then bring them
together. Change one thing, refit, look at the residuals again.
:::

The procedure is carried by the residuals: in a well specified model they
should show no trend and no large outliers.

## An age effect on F

```{r}
#| eval: false
fit02 <- sca(stk, idx,
             fmodel  = ~factor(age),   # <- was ~1
             qmodel  = list(~1),
             srmodel = ~1, vmodel = list(~1, ~1), n1model = ~1)
```

`factor(age)` means **as many parameters as ages**, each one
independent of the others.

The catch-at-age residuals become less staggered, reflecting the age effect.
The year trends are still there, but the average residual **by age** is now
much closer to 0.

## Age effect on F — residuals by year

```{r}
#| echo: false
#| fig-height: 3.0
plot(res02)
```

## Age effect on F — residuals by age

```{r}
#| echo: false
#| fig-height: 3.0
plot(res02, auxline = "l", by = "age")
```

## An age effect on q instead

Now add the age effect to the catchability submodel, and **remove** it from
the catch.

```{r}
#| eval: false
fit03 <- sca(stk, idx,
             fmodel  = ~1,
             qmodel  = list(~factor(age)),   # <- was ~1
             srmodel = ~1, vmodel = list(~1, ~1), n1model = ~1)
```

## Age effect on q — residuals by year

```{r}
#| echo: false
#| fig-height: 3.0
plot(res03)
```

## Age effect on q — residuals by age

```{r}
#| echo: false
#| fig-height: 3.0
plot(res03, auxline = "l", by = "age")
```

## Both together

```{r}
#| eval: false
fit04 <- sca(stk, idx,
             fmodel = ~factor(age), qmodel = list(~factor(age)),
             srmodel = ~1, vmodel = list(~1, ~1), n1model = ~1)
```

Finally the two effects are brought together.

## Both effects — residuals by year

```{r}
#| echo: false
#| fig-height: 3.0
plot(res04)
```

## Both effects — residuals by age

```{r}
#| echo: false
#| fig-height: 3.0
plot(res04, auxline = "l", by = "age")
```

## Now look at what it estimated

```{r}
#| echo: false
s4 <- stk + fit04
c(fbar = max(c(fbar(s4))), ssb = max(c(ssb(s4))))
```

::: {.callout-warning title="Converged, clean residuals, and nonsense"}
$\bar{F} \approx 5\times10^{-15}$ and SSB $\approx 3\times10^{17}$ tonnes:
no fishing at all, on an impossibly heavy stock.
:::

Residuals tell you whether the model reproduces the **data**. They cannot
tell you whether the answer is **possible**. Always look at $\bar{F}$, SSB
and recruitment as well.

Adding a year effect to $F$ — next module — brings $\bar{F}$ back to 0.6–2.8.

## Summary

- Explore the age effect on $F$ and on $q$ **separately**, then together
- `factor(age)` = one free parameter per age, independent of its neighbours
- On $F$: the catch residuals are less staggered and centre on 0 by age
- The year trends remain — that is Module 2.4
- **Check $\bar{F}$ and SSB before you believe any residual plot**

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M2.3_age_effects.R`.

1. Fit `fit02`, `fit03` and `fit04`
2. Plot the residuals of each **by year and by age**
3. Compare the fitted stocks with `FLStocks()`
:::
