---
title: "Reading residuals"
subtitle: "a4a Training Course — Module 2.2"
author: "Danai Mantopoulou Palouka & Alessandro Ligas"
date: today
format:
  # ---- standalone PDF: both the teaching slides and the handout -----------
  beamer:
    theme: metropolis
    themeoptions:
      - progressbar=frametitle
      - numbering=fraction
      - block=fill
    aspectratio: 169
    classoption: [t]
    pdf-engine: xelatex
    include-in-header: ../../_assets/a4a-beamer.tex
    highlight-style: github
    code-block-font-size: \scriptsize
    # beamer frames do not scroll — keep figures short so they fit.
    # Override per chunk with #| fig-height: when a plot needs more room.
    fig-width: 8
    fig-height: 2.8
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(FLCore)
library(FLa4a)
library(ggplotFL)
library(lattice)
library(latticeExtra)
lattice.options(default.args = list(as.table = TRUE))

data(mut09)
data(mut09.idx)
stk <- mut09
idx <- mut09.idx

fit01 <- sca(stk, idx,
             fmodel  = ~1,
             qmodel  = list(~1),
             srmodel = ~1,
             vmodel  = list(~1, ~1),
             n1model = ~1)

res <- residuals(fit01, stk, idx)
```

## Why this module exists

::: {.callout-warning title="Residuals are your main tool to manage your fit"}
If you cannot read them, you
cannot configure the model.
:::

We stay on the mean model from Module 2.1 — a fit we already know is bad —
and learn what each tool can show us.

## Three kinds of residual

| Type | Divided by | 
|:--|:--|
| `deviances` | nothing — the raw log difference |
| `standardized` | the **residual** standard deviation | 
| `pearson` | the model's **predicted** square root variance | 

With `vmodel = ~1` the variance is one constant, so pearson and standardized
look alike.

## Each tool answers a different question

| Tool | Question |
|:--|:--|
| `plot(res)` | does the fit drift through **time**? |
| `plot(res, by="age")` | is it off at particular **ages**? |
| `bubbles(res)` | same but with bubbles :-) |
| `qqmath(res)` | is the **error distribution** right? |
| `plot(fit, stk)` & `plot(fit, idx)`| check differences in patterns |
| `fitSumm(fit)` | number of parameters VS observations |

Look by year **and** by age, every time. A pattern invisible in one view is
often obvious in the other.

## Rows, columns, diagonals

```{r}
#| fig-height: 2.9
bubbles(res)
```


## Does it reproduce the catch? {.shrink}

Everything so far was **at age**. This asks the blunter question, on
aggregated catch in weight.

```{r}
#| fig-height: 2.2
cd <- computeCatchDiagnostics(fit01, stk)
plot(cd, type = "prediction", probs = c(0.025, 0.975))
```

The observed catch sits well below the predicted interval — the mean model
does not reproduce it at all.

::: {.notes}
plot(cd) on its own gives the full residual panel — a square multi-panel
figure, better shown live than on a slide. The message here is that a model
can look tidy at age and still miss total catch, which is the number
everyone in the room recognises.
:::

## Two cautions before you start

::: {.callout-warning title="Check these every time"}
`convergence` must be **0** in `fitSumm()`. Anything else and the rest of
the output is meaningless.

`AIC` and `BIC` only compare models fitted to the **same data** — never
across stocks or across different year/age ranges.
:::

Now go and look at a bad model properly.

## Hands-on

::: {.callout-tip title="Exercise"}
Open `handson_M2.2_residuals.R`.

1. Compute all three residual types and compare their spread
2. Work through `plot`, `bubbles`, `qqmath`, `plot(fit, stk)`
3. Build the catch diagnostics
4. Decide, from the evidence, what you would change first

Then try it: refit with an age effect on F and see whether the thing you
expected to improve actually did.
:::
